---
title: "PPI → AKI after ICI — Analysis"
subtitle: "Time-varying Cox model, 730-day horizon"
author: "Tianqi Ouyang"
date: "`r Sys.Date()`"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
## About this page
Renders **aggregate results only** from `cache/*.rds` built by
[`data_management.qmd`](data_management.qmd). No patient-level row reaches the
output; cells below 11 are suppressed. Follow-up horizon is 730 days.
:::

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr); library(survival); library(broom)
  library(gtsummary); library(ggplot2); library(knitr)
})
source("R/ppi_functions.R")
have <- all(file.exists(file.path("cache", paste0(c("qc","pt","cp","cp_ckd","cfg"), ".rds"))))
```

```{r}
#| label: load
#| include: false
if (!have) knitr::knit_exit()
QC     <- readRDS("cache/qc.rds")
pt     <- readRDS("cache/pt.rds")
cp     <- readRDS("cache/cp.rds")
cp_ckd <- readRDS("cache/cp_ckd.rds")
CFG    <- readRDS("cache/cfg.rds")

## Human-readable factor labels used by both tables. dm/htn/male are stored 0/1.
label_covars <- function(df) {
  df %>% mutate(
    Sex          = factor(if_else(male == 1, "Male", "Female"), c("Female", "Male")),
    Race         = factor(Race, c("White", "Black", "Asian", "Other/Unknown")),
    Diabetes     = factor(if_else(dm == 1, "Yes", "No"), c("No", "Yes")),
    Hypertension = factor(if_else(htn == 1, "Yes", "No"), c("No", "Yes"))
  )
}
```

```{r}
#| label: header
#| echo: false
#| output: asis
c0 <- QC$cohort
## Keep this ASCII only. An R cat()/sprintf() under a non-UTF-8 render locale
## turns any multibyte character (e.g. a middot separator) into a literal
## "<U+00B7>" in the output, so no such characters are used here.
cat(sprintf(paste0(
  "The cohort is **%s** new-user patients. **%s** start a PPI during follow-up ",
  "and **%s** never do. There are **%s** AKI events over %s person-years; ",
  "median follow-up is **%s** days.\n\n"),
  format(c0$n, big.mark=","), format(c0$ppi_initiators, big.mark=","),
  format(c0$never_ppi, big.mark=","), format(c0$aki_events, big.mark=","),
  format(c0$person_years, big.mark=","), c0$median_fu_days))
```

# 1 Cohort

**How the 8,911 analytic cohort is reached:**

```{r}
#| label: tbl-ledger
#| echo: false
kable(QC$ledger)
```

**Person-time and crude AKI rate by time-varying PPI state:**

```{r}
#| label: tbl-person-time
#| echo: false
kable(QC$person_time)
```

# 2 Table 1 — baseline characteristics

Baseline covariates for the whole cohort and by whether the patient starts a PPI
during follow-up. PPI initiation is **time-varying** in the model; this split is
descriptive only (it groups patients by an exposure that, for initiators, is
acquired after baseline). Continuous variables are mean (SD); categorical are
n (%).

```{r}
#| label: table1
#| echo: false
t1_data <- pt %>%
  label_covars() %>%
  transmute(
    `PPI during follow-up` = factor(if_else(ppi_initiator == 1, "Started PPI", "Never PPI"),
                                    c("Never PPI", "Started PPI")),
    `Age at ICI start, years`        = age_ici,
    Sex, Race, Diabetes, Hypertension,
    `Baseline eGFR, mL/min/1.73 m2`  = eGFR_CRE_baseline
  )

tbl_summary(
  t1_data,
  by = `PPI during follow-up`,
  statistic = list(all_continuous() ~ "{mean} ({sd})",
                   all_categorical() ~ "{n} ({p}%)"),
  digits = all_continuous() ~ 1
) %>%
  add_overall(last = FALSE) %>%
  modify_header(label ~ "**Characteristic**") %>%
  modify_caption("**Table 1. Baseline characteristics of the 8,911 new-user cohort.**")
```

# 3 Time-varying Cox model

Every patient is unexposed at ICI start and becomes exposed only from the day
they fill a PPI, so pre-PPI follow-up counts as unexposed person-time. The model
is `Surv(tstart, tstop, AKI) ~ PPI (+ covariates)`.

```{r}
#| label: cox-fit
#| echo: false
cp_lab <- cp %>%
  label_covars() %>%
  mutate(PPI = factor(if_else(ppi == 1, "On PPI", "Not on PPI"), c("Not on PPI", "On PPI")))

m_crude <- coxph(Surv(tstart, tstop, event) ~ PPI, data = cp_lab)
m_adj   <- coxph(Surv(tstart, tstop, event) ~ PPI + age_ici + Sex + Race +
                   Diabetes + Hypertension + eGFR_CRE_baseline, data = cp_lab)

n_pt <- dplyr::n_distinct(cp_lab$id); n_ev <- sum(cp_lab$event)
```

**The PPI effect (On PPI vs Not on PPI), unadjusted and adjusted:**

```{r}
#| label: tbl-ppi-effect
#| echo: false
bind_rows(
  broom::tidy(m_crude, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("PPI", term)) %>% mutate(model = "Unadjusted"),
  broom::tidy(m_adj, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("PPI", term)) %>% mutate(model = "Adjusted")
) %>%
  transmute(Model = model,
            `HR for AKI (95% CI)` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
            p = ifelse(p.value < .001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable()
```

Full adjusted model — every covariate:

```{r}
#| label: tbl-cox-full
#| echo: false
tbl_regression(m_adj, exponentiate = TRUE,
               label = list(age_ici ~ "Age at ICI start (per year)",
                            eGFR_CRE_baseline ~ "Baseline eGFR (per mL/min/1.73 m2)")) %>%
  modify_header(estimate ~ "**HR**") %>%
  modify_caption(sprintf(
    "**Table 2. Multivariable time-varying Cox model for AKI.** %s patients, %s AKI events over %s person-years.",
    format(n_pt, big.mark = ","), format(n_ev, big.mark = ","),
    format(QC$cohort$person_years, big.mark = ",")))
```

```{r}
#| label: ph-check
#| echo: false
#| output: asis
z  <- cox.zph(m_adj)
zt <- as.data.frame(z$table)["PPI", , drop = FALSE]
cat(sprintf("_Proportional-hazards check for the PPI term (cox.zph): p = %s._\n\n",
            ifelse(zt$p < .001, "<0.001", sprintf("%.3f", zt$p))))
```

# 4 Cumulative incidence of AKI

Cumulative incidence of AKI over two years for the whole cohort (death and end
of creatinine follow-up treated as censoring).

```{r}
#| label: fig-cuminc
#| echo: false
#| fig-height: 4.2
#| fig-cap: "Cumulative incidence of AKI within 730 days of ICI start (8,911 new-users)."
km <- survfit(Surv(fu_days, ev) ~ 1, data = pt)
data.frame(time = km$time, ci = 1 - km$surv) |>
  ggplot(aes(time, ci)) +
  geom_step(color = "#0B6E6D", linewidth = 0.9) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  labs(x = "Days since ICI start", y = "Cumulative incidence of AKI") +
  theme_minimal(base_size = 12)
```

# 5 CKD composite outcome (Kavita definition)

Second outcome, using the **same CKD composite as the Kavita ICI-CKD project**:
ESRD / kidney transplant after ICI, sustained (>=90 day) eGFR < 10, incident CKD
(baseline eGFR >= 60 with a sustained > 30% drop to eGFR < 60), or CKD
progression (baseline eGFR < 60 sustained). Exposure and covariates are exactly
as in the AKI model above; only the outcome and its follow-up clock change.

**Components contributing to the composite (new-user cohort):**

```{r}
#| label: tbl-ckd-components
#| echo: false
kable(QC$ckd_components)
```

**Person-time and crude CKD rate by time-varying PPI state:**

```{r}
#| label: tbl-ckd-person-time
#| echo: false
kable(QC$person_time_ckd)
```

```{r}
#| label: ckd-cox
#| echo: false
cpc <- cp_ckd %>%
  label_covars() %>%
  mutate(PPI = factor(if_else(ppi == 1, "On PPI", "Not on PPI"), c("Not on PPI", "On PPI")))

mc_crude <- coxph(Surv(tstart, tstop, event) ~ PPI, data = cpc)
mc_adj   <- coxph(Surv(tstart, tstop, event) ~ PPI + age_ici + Sex + Race +
                    Diabetes + Hypertension + eGFR_CRE_baseline, data = cpc)
nc_pt <- dplyr::n_distinct(cpc$id); nc_ev <- sum(cpc$event)
```

**The PPI effect on CKD composite (On PPI vs Not on PPI), unadjusted and adjusted:**

```{r}
#| label: tbl-ckd-effect
#| echo: false
bind_rows(
  broom::tidy(mc_crude, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("PPI", term)) %>% mutate(model = "Unadjusted"),
  broom::tidy(mc_adj, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("PPI", term)) %>% mutate(model = "Adjusted")
) %>%
  transmute(Model = model,
            `HR for CKD composite (95% CI)` =
              sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
            p = ifelse(p.value < .001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable()
```

Full adjusted model — every covariate:

```{r}
#| label: tbl-ckd-cox-full
#| echo: false
tbl_regression(mc_adj, exponentiate = TRUE,
               label = list(age_ici ~ "Age at ICI start (per year)",
                            eGFR_CRE_baseline ~ "Baseline eGFR (per mL/min/1.73 m2)")) %>%
  modify_header(estimate ~ "**HR**") %>%
  modify_caption(sprintf(
    "**Table 3. Multivariable time-varying Cox model for the CKD composite.** %s patients, %s CKD events over %s person-years.",
    format(nc_pt, big.mark = ","), format(nc_ev, big.mark = ","),
    format(QC$cohort_ckd$person_years, big.mark = ",")))
```

```{r}
#| label: ckd-ph-check
#| echo: false
#| output: asis
zc  <- cox.zph(mc_adj)
zct <- as.data.frame(zc$table)["PPI", , drop = FALSE]
cat(sprintf("_Proportional-hazards check for the PPI term (cox.zph): p = %s._\n\n",
            ifelse(zct$p < .001, "<0.001", sprintf("%.3f", zct$p))))
```

**Cumulative incidence of the CKD composite:**

```{r}
#| label: fig-ckd-cuminc
#| echo: false
#| fig-height: 4.2
#| fig-cap: "Cumulative incidence of the CKD composite within 730 days of ICI start (8,911 new-users; death and end of creatinine follow-up treated as censoring)."
kmc <- survfit(Surv(fu_ckd, ev_ckd) ~ 1, data = pt)
data.frame(time = kmc$time, ci = 1 - kmc$surv) |>
  ggplot(aes(time, ci)) +
  geom_step(color = "#8a4b0f", linewidth = 0.9) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, NA)) +
  labs(x = "Days since ICI start", y = "Cumulative incidence of CKD composite") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: session
#| echo: false
#| output: asis
cat("\n---\n\n*Built from `cache/qc.rds` created ",
    format(QC$built_at, "%Y-%m-%d %H:%M"),
    ". Aggregate results only; cells below ", CFG$SMALL_CELL, " suppressed.*\n", sep = "")
```
