---
title: "PPI → AKI after ICI — Analysis"
subtitle: "Time-varying Cox models, 730-day horizon"
author: "Tianqi Ouyang"
date: "`r Sys.Date()`"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
## About this page

Reads `cache/*.rds` built by [`data_management.qmd`](data_management.qmd) and
renders **aggregate results only** — model coefficients, person-time totals and
rates. No patient-level row reaches the output, and cells below 11 are
suppressed. Helper functions come from [`R/ppi_functions.R`](R/ppi_functions.R).

Follow-up horizon is **730 days (2 years)**. The two reference summaries were
computed at 548 days, so nothing here reproduces them exactly.
:::

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(purrr); library(stringr)
  library(survival); library(broom); library(ggplot2); library(knitr)
})
source("R/ppi_functions.R")

CACHE_OK <- all(file.exists(file.path("cache",
  paste0(c("qc","pt","cp","cp_base","dens","model_vars","cfg"), ".rds"))))
```

```{r}
#| label: load-cache
#| include: false
if (!CACHE_OK) knitr::knit_exit()

QC         <- readRDS("cache/qc.rds")
pt         <- readRDS("cache/pt.rds")
cp         <- readRDS("cache/cp.rds")
cp_base    <- readRDS("cache/cp_base.rds")
dens       <- readRDS("cache/dens.rds")
MODEL_VARS <- readRDS("cache/model_vars.rds")
CFG        <- readRDS("cache/cfg.rds")

## Shared model machinery -------------------------------------------------------
ADJ <- paste(MODEL_VARS, collapse = " + ")

#' Fit a counting-process Cox model and return the exponentiated terms of
#' interest. na.action is set explicitly: bug A10(1) is that the default
#' silently drops INTERVALS, not patients. Missingness was already resolved at
#' the patient level in data_management.qmd §9, so na.fail is a real assertion
#' rather than a formality.
fit_cp <- function(data, rhs, keep = "^ppi", label = NA_character_,
                   outcome = "Surv(tstart, tstop, event)") {
  f <- stats::as.formula(paste(outcome, "~", rhs))
  m <- coxph(f, data = data, na.action = stats::na.fail)
  list(
    label = label, model = m,
    tidy = broom::tidy(m, exponentiate = TRUE, conf.int = TRUE) %>%
      filter(grepl(keep, term)) %>%
      transmute(model = label, term,
                HR = estimate, lo = conf.low, hi = conf.high, p = p.value),
    n_id = length(unique(data$id)), n_rows = nrow(data),
    n_events = sum(data$event),
    py = sum(data$tstop - data$tstart) / 365.25
  )
}

hr_txt <- function(d) sprintf("%.2f (%.2f-%.2f)", d$HR, d$lo, d$hi)

show_hr <- function(..., caption = NULL, digits = 3) {
  bind_rows(lapply(list(...), `[[`, "tidy")) %>%
    mutate(`HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, lo, hi),
           p = ifelse(p < .001, "<0.001", sprintf("%.3f", p))) %>%
    select(model, term, `HR (95% CI)`, p) %>%
    kable(caption = caption)
}
```

```{r}
#| label: cohort-header
#| echo: false
#| output: asis
i <- QC$initiators
cat(sprintf(paste0(
  "**Cohort** %s new-user patients · **%s** PPI initiators under observation ",
  "· **%s** never · **%s** AKI events · %s person-years ",
  "· median follow-up %s days · %s%% still under creatinine surveillance at 730 d\n\n"),
  format(i$n_cohort, big.mark=","), format(i$n_initiator, big.mark=","),
  format(i$n_never, big.mark=","), format(i$n_events, big.mark=","),
  format(round(i$py), big.mark=","), i$median_fu, i$pct_followed_730))
```

---

# 1 Cohort & person-time

```{r}
#| label: tbl-ledger
#| echo: false
kable(QC$ledger, caption = "CONSORT ledger. Every step executed and asserted in code (bug A13).")
kable(QC$ppi_prev,
      caption = paste0("PPI prevalence by definition, denominator ",
                       format(QC$ppi_prev_denom, big.mark = ","),
                       ". The unguarded med_list() reported 71.3% as baseline (bug A4)."))
kable(QC$person_time, caption = "Person-time by time-varying exposure state")
```

## 1.1 The two initiator definitions (bug A11)

The reference summaries used three initiator counts interchangeably — 2,951,
3,058 and 3,341. The gap is the **observation window**: whether a PPI that
arrives after the patient's follow-up has already ended still counts.

```{r}
#| label: tbl-initiator-defs
#| echo: false
tibble(
  definition = c(
    "Initiation while UNDER OBSERVATION  (model exposure, ppi_initiator)",
    "Any post-ICI PPI within 730 d, ignoring follow-up end (ppi_ever_post)",
    "Difference: PPI arrives after AKI / death / last creatinine"),
  n = c(QC$initiators$n_initiator, QC$initiators$n_ever_post_full,
        QC$initiators$n_ever_post_full - QC$initiators$n_initiator)
) %>% kable(caption = "Both definitions are needed: the first drives every model, the second drives the mirror window and the immortal-time diagnostic.")
```

---

# 2 The AKI outcome was repaired first (bug A1)

`create_aki()` computed its KDIGO lookback windows **after** discarding every
pre-ICI creatinine, so neither comparator could see a baseline value and AKI was
structurally undetectable at each patient's first post-ICI draw. The table below
is the same function run both ways on the same input.

```{r}
#| label: tbl-aki-fix
#| echo: false
f <- QC$aki_fix
tibble::tribble(
  ~Quantity,                                              ~Original, ~Fixed,
  "Patients with AKI (parent cohort, ascertainable)",      f$n_aki_buggy, f$n_aki_fixed,
  "Events in days 0-90 after ICI",                        f$n_events_d0_90_buggy, f$n_events_d0_90_fixed,
  "Patients with aki_days == 1",                           f$n_day1_buggy, f$n_day1_fixed,
  "Earliest detectable aki_days",                          f$min_day_buggy, f$min_day_fixed
) %>% kable(caption = "Effect of the A1 fix. aki_days == 1 is impossible under the original ordering -- that is the bug's fingerprint, not biology.")

tibble(
  change = c("Gained an AKI", "Lost an AKI", "Kept AKI, date moved",
             "Used pre_CRE_180days as KDIGO baseline"),
  n = c(f$n_gained, f$n_lost, f$n_date_moved, f$n_fallback_patients)
) %>% kable(caption = "Patient-level movement. Events can also be LOST: adding pre-ICI values to the 90-day window can raise the median and pull cre/median below 1.5.")
```

::: {.callout-warning}
## This affects the published paper, not only this analysis
The same broken ordering is in the ICI-CKD pipeline at `data_management.qmd:494`
and in `Final cohort/aki.Rmd:82`, which produced `aki_0120.xlsx`. Any result
built on that file has an undercounted AKI outcome, concentrated in the first
90 days.
:::

---

# 3 M0 — the naive time-fixed model, and why it is wrong

Classifying a patient as "PPI-exposed" from day 0 because they started a PPI on
day 120 credits 120 drug-free days to the drug (bug A5).

```{r}
#| label: tbl-immortal
#| echo: false
it <- QC$immortal_time
tibble(
  quantity = c("Naive 'ever initiated' exposed patients",
               "Their AKI events",
               "  ... occurring BEFORE the first PPI",
               "  ... on the SAME DAY as the first PPI",
               "  ... AFTER the first PPI",
               "% of naive exposed person-days that precede the drug"),
  value = c(it$n_naive_exposed, it$n_events_naive_exposed,
            it$n_events_before_ppi, it$n_events_same_day,
            it$n_events_after_ppi, it$pct_predrug_persondays)
) %>% kable(caption = "Immortal time bias, measured. A model that calls these events 'exposed' is uninterpretable whichever way it errs.")
```

```{r}
#| label: m0-naive
naive_pt <- pt %>%
  mutate(ppi_naive = ppi_ever_post) %>%
  select(id = EMPI, fu_days, ev, ppi_naive, all_of(MODEL_VARS))

m0 <- coxph(as.formula(paste("Surv(fu_days, ev) ~ ppi_naive +", ADJ)),
            data = naive_pt, na.action = na.fail)
M0 <- list(label = "M0 naive time-fixed",
           tidy = broom::tidy(m0, exponentiate = TRUE, conf.int = TRUE) %>%
             filter(term == "ppi_naive") %>%
             transmute(model = "M0 naive time-fixed", term, HR = estimate,
                       lo = conf.low, hi = conf.high, p = p.value))
```

::: {.callout-important}
`ppi_naive` above uses `ppi_ever_post` — the full-record definition. Using the
model's truncated `ppi_day` would make pre-PPI events impossible **by
construction** and hide the very bias M0 exists to demonstrate.
:::

---

# 4 M1-M4 — time-varying Cox

One row per covariate interval; `Surv(tstart, tstop, event)` replaces
`Surv(time, event)`. **M1 is the primary model as specified.** M4 adds the
pre-specified 90-day induction period, with that person-time deleted from the
risk set rather than reassigned to the comparator.

```{r}
#| label: m1-m4
M1 <- fit_cp(cp$lag0,  paste("ppi +", ADJ),                      label = "M1 time-varying, lag 0")
M2 <- fit_cp(cp$lag0,  paste("ppi + ster +", ADJ),               label = "M2 + time-varying steroid")
M3 <- fit_cp(cp$lag0,  paste("ppi + ster + labdens +", ADJ),     label = "M3 + draw density")
M4 <- fit_cp(cp$lag90, paste("ppi + ster + labdens +", ADJ),     label = "M4 + 90-day induction")

show_hr(M1, M2, M3, M4, caption = "Primary time-varying models. Adjusted for age, sex, race, ICI year, DM, HTN, CAD, cirrhosis, diuretic, ACE/ARB, statin, smoking, baseline steroid, baseline eGFR/Hgb/albumin (+ missing indicators), 8 chemo flags, cancer group. esrd_kt omitted: constant zero by construction.")
```

```{r}
#| label: tbl-steroid-term
#| echo: false
broom::tidy(M3$model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term %in% c("ppi", "ster", "labdens")) %>%
  transmute(term, `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
            p = ifelse(p.value < .001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "M3 terms of interest. The steroid and draw-density coefficients are themselves part of the argument (bugs A7, A8).")
```

## 4.1 Proportional hazards

A PH violation is expected here and is informative rather than fatal: it is the
same decaying hazard that §6 quantifies directly. The correct response is
time-stratified estimates, not one number.

```{r}
#| label: cox-zph
#| echo: false
zph <- cox.zph(M3$model)
as.data.frame(zph$table) %>%
  tibble::rownames_to_column("term") %>%
  filter(term %in% c("ppi", "ster", "labdens", "GLOBAL")) %>%
  mutate(p = ifelse(p < .001, "<0.001", sprintf("%.3f", p))) %>%
  select(term, chisq, df, p) %>%
  kable(digits = 2, caption = "Schoenfeld residual test (cox.zph) on M3")
```

---

# 5 M5 — the induction ladder

Push the exposure start later and later, deleting the induction window from the
risk set each time. A pharmacological effect would plateau.

```{r}
#| label: m5-ladder
ladder <- purrr::imap_dfr(cp[paste0("lag", CFG$LAGS)], function(d, nm) {
  lg <- as.integer(sub("lag", "", nm))
  f  <- fit_cp(d, paste("ppi + ster + labdens +", ADJ),
               label = sprintf("induction %d d", lg))
  f$tidy %>% mutate(lag = lg, events = f$n_events, py = round(f$py))
}) %>% arrange(lag)

ladder %>%
  transmute(`Induction period` = ifelse(lag == 0, "none (day 0)", paste(lag, "days")),
            events, `person-years` = py,
            `Adjusted HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, lo, hi)) %>%
  kable(caption = "Every extra day of separation between the prescription and the risk window removes more of the effect.")
```

```{r}
#| label: fig-ladder
#| echo: false
#| fig-height: 3.2
#| fig-cap: "Induction ladder. Monotone decay with no plateau is not the signature of a drug effect."
ggplot(ladder, aes(x = factor(lag), y = HR)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(ymin = lo, ymax = hi), size = .7) +
  scale_y_continuous(trans = "log", breaks = c(0.8,1,1.5,2,2.5,3)) +
  labs(x = "Induction period (days)", y = "Adjusted HR for AKI (log scale)") +
  theme_minimal(base_size = 12)
```

---

# 6 M6 — hazard by time since initiation

Split the exposed person-time by how long the patient has been on the drug.

```{r}
#| label: m6-bands
BAND_CUTS <- c(0, 7, 30, 90, 365)
band_base <- cp_base %>%
  mutate(!!!setNames(
    lapply(BAND_CUTS, function(k) quote(ppi_day)), paste0("d", BAND_CUTS)))
for (k in BAND_CUTS) band_base[[paste0("d", k)]] <- band_base$ppi_day + k

cp_band <- build_counting_process(
  band_base,
  switches = c(setNames(as.list(paste0("d", BAND_CUTS)), paste0("on", BAND_CUTS)),
               list(ster = "ster_day")),
  dens = dens
)

cp_band <- cp_band %>%
  mutate(ppi_band = factor(
    rowSums(across(all_of(paste0("on", BAND_CUTS)))),
    levels = 0:5,
    labels = c("unexposed", "day 0-7", "day 8-30", "day 31-90",
               "day 91-365", "beyond 1 year")))

M6 <- fit_cp(cp_band, paste("ppi_band + ster + labdens +", ADJ),
             keep = "^ppi_band", label = "M6 AKI")
```

```{r}
#| label: m6-death
## Same banding on the death clock -- the negative-control outcome (M7).
band_death <- cp$death %>% select(id) %>% distinct()
bd_base <- pt %>%
  transmute(id = EMPI, fu_days = fu_days_death, ev = ev_death,
            ppi_day = if_else(!is.na(ppi_day_full) & ppi_day_full <= fu_days_death,
                              ppi_day_full, NA_real_),
            ster_day = as.numeric(ster_first_post_date - ICI_Dose_Date)) %>%
  mutate(ster_day = if_else(!is.na(ster_day) & ster_day <= fu_days, ster_day, NA_real_)) %>%
  bind_cols(pt %>% select(all_of(MODEL_VARS)))
for (k in BAND_CUTS) bd_base[[paste0("d", k)]] <- bd_base$ppi_day + k

cp_band_death <- build_counting_process(
  bd_base,
  switches = c(setNames(as.list(paste0("d", BAND_CUTS)), paste0("on", BAND_CUTS)),
               list(ster = "ster_day")),
  dens = dens
) %>%
  mutate(ppi_band = factor(rowSums(across(all_of(paste0("on", BAND_CUTS)))),
                           levels = 0:5,
                           labels = c("unexposed","day 0-7","day 8-30","day 31-90",
                                      "day 91-365","beyond 1 year")))

M7 <- fit_cp(cp_band_death, paste("ppi_band + ster + labdens +", ADJ),
             keep = "^ppi_band", label = "M7 death")
```

```{r}
#| label: tbl-bands
#| echo: false
rates <- cp_band %>%
  group_by(ppi_band) %>%
  summarise(py = sum(tstop - tstart) / 365.25, n = sum(event), .groups = "drop") %>%
  mutate(`AKI /100py` = round(100 * n / py, 1),
         py = round(py), events = suppress_small_cells(n, CFG$SMALL_CELL)) %>%
  select(ppi_band, `person-yrs` = py, events, `AKI /100py`)

aki_hr <- M6$tidy %>% mutate(ppi_band = sub("^ppi_band", "", term)) %>%
  transmute(ppi_band, `AKI HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, lo, hi))
dth_hr <- M7$tidy %>% mutate(ppi_band = sub("^ppi_band", "", term)) %>%
  transmute(ppi_band, `Death HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, lo, hi))

rates %>%
  left_join(aki_hr, by = "ppi_band") %>%
  left_join(dth_hr, by = "ppi_band") %>%
  mutate(across(everything(), ~ tidyr::replace_na(as.character(.), "1.00 (ref)"))) %>%
  kable(caption = "Hazard by time since first PPI, with death as a co-primary negative control. Creatinine sampling can manufacture an AKI; it cannot manufacture a death.")
```

::: {.callout-note}
## What to read off this table
Two things. First, whether the AKI hazard under **continuing** exposure decays
toward the unexposed reference — a nephrotoxin does not get safer the longer it
is taken. Second, whether the **death** hazard exceeds the AKI hazard in the
same person-time — an exposure that raises all-cause mortality more than it
raises kidney injury is a severity marker, not a nephrotoxin.
:::

---

# 7 The mirror window (bug A6)

The decisive descriptive object. AKI rate by day **relative to the first PPI**,
looking backward as well as forward. If a PPI caused AKI, risk would be flat
before the prescription and raised after.

```{r}
#| label: tbl-mirror
#| echo: false
QC$mirror %>%
  transmute(Band = band, `person-days` = person_days,
            events = events_shown, `AKI /1,000 person-days` = rate_per_1000_pd) %>%
  kable(caption = paste0("Among ", format(QC$mirror_n_patients, big.mark=","),
    " patients with a post-ICI PPI. Computed on the REPAIRED AKI outcome and on ",
    "full-record dates; using follow-up-truncated dates would force every ",
    "pre-prescription band to zero."))
```

```{r}
#| label: fig-mirror
#| echo: false
#| fig-height: 4.2
#| fig-cap: "Mirror window. A drug cannot raise risk in the week before it is given, so a symmetric spike centred on day 0 indicates that the PPI order and the creatinine rise are two readouts of one acute care episode."
QC$mirror %>%
  mutate(side = case_when(hi < 0 ~ "before the prescription",
                          lo > 0 ~ "after the prescription",
                          TRUE   ~ "prescription day")) %>%
  ggplot(aes(band, rate_per_1000_pd, fill = side)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.2f", rate_per_1000_pd)), vjust = -0.4, size = 3.1) +
  scale_fill_manual(values = c("before the prescription" = "#3E9695",
                               "prescription day"        = "#B84A34",
                               "after the prescription"  = "#C8862A")) +
  labs(x = "Days relative to first post-ICI PPI", y = "AKI per 1,000 person-days",
       fill = NULL) +
  expand_limits(y = max(QC$mirror$rate_per_1000_pd) * 1.12) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: mirror-symmetry
#| echo: false
#| output: asis
w_before <- QC$mirror$rate_per_1000_pd[QC$mirror$band == "-7..-1"]
w_after  <- QC$mirror$rate_per_1000_pd[QC$mirror$band == "+1..+7"]
d0       <- QC$mirror$rate_per_1000_pd[QC$mirror$band == "day 0"]
cat(sprintf(paste0("Week **before** the prescription: **%.2f** per 1,000 person-days. ",
  "Week **after**: **%.2f**. Ratio before/after: **%.2f**. ",
  "Prescription day itself: **%.2f**.\n\n"), w_before, w_after, w_before/w_after, d0))
```

---

# 8 M8 — H2-blocker active comparator

H2RAs treat the same indication, arrive in the same acute encounters, and lack
the interstitial-nephritis mechanism. A null PPI-vs-H2RA contrast is the
cleanest confirmation that the signal is the encounter rather than the pill.

```{r}
#| label: m8-h2ra
h2_base <- cp_base %>%
  left_join(pt %>% select(id = EMPI, h2ra_day2 = h2ra_day, h2ra_pre), by = "id") %>%
  filter(h2ra_pre == 0)          # new user of BOTH classes

cp_h2 <- build_counting_process(
  h2_base, switches = list(ppi = "ppi_day", h2ra = "h2ra_day2", ster = "ster_day"),
  dens = dens)

M8 <- fit_cp(cp_h2, paste("ppi + h2ra + ster + labdens +", ADJ),
             keep = "^(ppi|h2ra)$", label = "M8 PPI vs H2RA")
show_hr(M8, caption = "Both exposures time-varying in one model, among patients with neither drug at baseline. The PPI-versus-H2RA contrast is the ratio of the two HRs.")
```

```{r}
#| label: m8-contrast
#| echo: false
#| output: asis
cf <- coef(M8$model); V <- vcov(M8$model)
if (all(c("ppi","h2ra") %in% names(cf))) {
  d  <- cf["ppi"] - cf["h2ra"]
  se <- sqrt(V["ppi","ppi"] + V["h2ra","h2ra"] - 2*V["ppi","h2ra"])
  cat(sprintf(paste0("**PPI versus H2RA** (ratio of hazard ratios): ",
    "**%.2f (%.2f-%.2f)**, Wald p = %s. New H2RA users available: %s.\n\n"),
    exp(d), exp(d-1.96*se), exp(d+1.96*se),
    ifelse(2*pnorm(-abs(d/se)) < .001, "<0.001", sprintf("%.3f", 2*pnorm(-abs(d/se)))),
    format(QC$initiators$n_h2ra_new_730, big.mark = ",")))
}
```

---

# 9 M9 — steroid initiation as a negative-control *exposure*

Among **never-PPI** patients, does starting a steroid reproduce the same decay
curve? If it does, the shape is a property of "a new drug was ordered", not of
PPI (bug A7).

```{r}
#| label: m9-steroid-nce
nce_base <- cp_base %>% filter(is.na(ppi_day))     # never-PPI only
nce <- purrr::map_dfr(c(0, 7, 30), function(lg) {
  d <- build_counting_process(nce_base, switches = list(ster = "ster_day"),
                              lag_days = lg, lag_on = if (lg > 0) "ster" else NULL,
                              dens = dens)
  fit_cp(d, paste("ster + labdens +", ADJ), keep = "^ster$",
         label = sprintf("steroid, induction %d d", lg))$tidy
})
nce %>% transmute(model, `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, lo, hi),
                  p = ifelse(p < .001, "<0.001", sprintf("%.3f", p))) %>%
  kable(caption = "Steroid initiation in never-PPI patients. A decay curve of the same shape means the pattern belongs to new-prescription timing, not to PPI.")
```

---

# 10 M11 — landmark analyses

A different estimator, not merely a different lag. At each landmark, restrict to
patients still at risk, fix exposure as *initiated by the landmark*, and follow
forward from there.

```{r}
#| label: m11-landmark
landmark <- purrr::map_dfr(c(90, 180, 365), function(L) {
  d <- pt %>%
    filter(fu_days > L) %>%
    mutate(exposed_by_L = as.integer(!is.na(ppi_day_full) & ppi_day_full <= L),
           t_start = 0, t_stop = fu_days - L, evL = ev)
  m <- coxph(as.formula(paste("Surv(t_stop, evL) ~ exposed_by_L +", ADJ)),
             data = d, na.action = na.fail)
  broom::tidy(m, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "exposed_by_L") %>%
    transmute(landmark = paste(L, "days"), n = nrow(d), events = sum(d$evL),
              `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
              p = ifelse(p.value < .001, "<0.001", sprintf("%.3f", p.value)))
})
kable(landmark, caption = "Landmark designs at 90 / 180 / 365 days.")
```

---

# 11 M10 — Fine-Gray with death as a competing risk

Initiators die more often than non-initiators, so a reviewer will ask.

Fine-Gray with a genuinely time-varying exposure is not well defined — the
subdistribution hazard is specified over a risk set that retains patients after
the competing event, so a covariate that switches during follow-up has no
consistent value there. Exposure is therefore fixed at the **90-day landmark**,
which is the estimator a reviewer expects and which stays interpretable.

```{r}
#| label: m10-finegray
L <- 90
fg_dat <- pt %>%
  filter(fu_days > L) %>%
  mutate(
    exposed_by_L = as.integer(!is.na(ppi_day_full) & ppi_day_full <= L),
    t   = pmin(fu_days, fu_days_death) - L,
    cr  = factor(case_when(ev == 1       ~ "aki",
                           ev_death == 1 ~ "death",
                           TRUE          ~ "censor"),
                 levels = c("censor", "aki", "death"))
  ) %>%
  filter(t > 0)

fg <- finegray(Surv(t, cr) ~ ., data = fg_dat, etype = "aki")
M10 <- coxph(as.formula(paste("Surv(fgstart, fgstop, fgstatus) ~ exposed_by_L +", ADJ)),
             data = fg, weights = fgwt, na.action = na.fail)

broom::tidy(M10, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term == "exposed_by_L") %>%
  transmute(model = "M10 Fine-Gray (90-d landmark exposure)",
            `sHR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
            p = ifelse(p.value < .001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "Subdistribution hazard for AKI with death as the competing event.")
```

```{r}
#| label: tbl-mortality
#| echo: false
pt %>%
  group_by(`PPI initiator` = ifelse(ppi_initiator == 1, "yes", "no")) %>%
  summarise(n = dplyr::n(), deaths = sum(ev_death),
            `% dead by 730 d` = round(100 * mean(ev_death), 1), .groups = "drop") %>%
  kable(caption = "Mortality by initiator status. The imbalance is why the competing-risk analysis is asked for, and part of why the exposure reads as a severity marker.")
```

---

# 12 M12 — outpatient-only initiators

One PPI order means "exposed forever", and inpatient IV pantoprazole is not
outpatient oral maintenance. `Route` is absent from this RPDR extract, but
`Inpatient_Outpatient` is present, so care setting can be modelled (bug A9).

```{r}
#| label: m12-outpatient
op_base <- cp_base %>% mutate(ppi_day = ppi_op_day)
cp_op <- build_counting_process(op_base, switches = list(ppi = "ppi_day", ster = "ster_day"),
                                dens = dens)
M12 <- fit_cp(cp_op, paste("ppi + ster + labdens +", ADJ),
              label = "M12 outpatient-only initiators")
show_hr(M12, caption = "Initiation defined only by a record flagged outpatient.")
```

```{r}
#| label: tbl-exposure-limits
#| echo: false
pt %>% filter(ppi_ever_post == 1) %>%
  summarise(
    `Initiators (full record)`      = dplyr::n(),
    `With an outpatient record`     = sum(!is.na(ppi_op_day)),
    `Single PPI record in follow-up`= sum(ppi_n_records == 1, na.rm = TRUE),
    `% single record`               = round(100 * mean(ppi_n_records == 1, na.rm = TRUE), 1)
  ) %>% t() %>% as.data.frame() %>% tibble::rownames_to_column(" ") %>%
  setNames(c(" ", "value")) %>%
  kable(caption = "Exposure resolution limits. Record count is NOT reported as dose-response (bug A9): in person-time strictly before any PPI, AKI incidence already rises with the number of records the patient will LATER accumulate, and a future prescription cannot cause a past AKI.")
```

---

# 13 All models together

```{r}
#| label: tbl-summary
#| echo: false
bind_rows(
  M0$tidy,
  M1$tidy, M2$tidy, M3$tidy, M4$tidy,
  ladder %>% filter(lag %in% c(30, 180)) %>%
    mutate(model = paste0("M5 induction ", lag, " d")) %>% select(names(M1$tidy)),
  M12$tidy
) %>%
  filter(term %in% c("ppi", "ppi_naive")) %>%
  transmute(Model = model, `HR for AKI (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, lo, hi),
            p = ifelse(p < .001, "<0.001", sprintf("%.3f", p))) %>%
  kable(caption = "Every model, same 8,911 patients and same repaired AKI outcome; only the handling of time changes.")
```

```{r}
#| label: fig-forest
#| echo: false
#| fig-height: 3.6
#| fig-cap: "The more the prescription is separated from the risk window, the more the association disappears, and it never plateaus."
bind_rows(M0$tidy, M1$tidy, M2$tidy, M3$tidy, M4$tidy, M12$tidy) %>%
  filter(term %in% c("ppi", "ppi_naive")) %>%
  mutate(model = factor(model, levels = rev(model))) %>%
  ggplot(aes(HR, model)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey40") +
  geom_pointrange(aes(xmin = lo, xmax = hi)) +
  scale_x_continuous(trans = "log", breaks = c(0.8,1,1.5,2,2.5,3,4)) +
  labs(x = "Hazard ratio for AKI (log scale)", y = NULL) +
  theme_minimal(base_size = 11)
```

---

# 14 Interpretation

```{r}
#| label: conclusion
#| echo: false
#| output: asis
p_hr <- M1$tidy; i_hr <- M4$tidy
b <- M6$tidy %>% filter(grepl("beyond", term))
dth <- M7$tidy %>% filter(grepl("91-365", term))
aki365 <- M6$tidy %>% filter(grepl("91-365", term))
cat(sprintf(paste0(
"- **M1, the specified primary time-varying model:** HR %s.\n",
"- **M4, with a 90-day induction period:** HR %s.\n",
"- **Beyond one year of continuing PPI:** HR %s for AKI.\n",
"- **Same person-time, days 91-365:** AKI HR %s versus **death** HR %s.\n",
"- **Mirror window:** %.2f per 1,000 person-days in the week before the ",
"prescription against %.2f in the week after (ratio %.2f), peaking at %.2f on ",
"the prescription day itself.\n\n"),
hr_txt(p_hr), hr_txt(i_hr), hr_txt(b), hr_txt(aki365), hr_txt(dth),
QC$mirror$rate_per_1000_pd[QC$mirror$band == "-7..-1"],
QC$mirror$rate_per_1000_pd[QC$mirror$band == "+1..+7"],
QC$mirror$rate_per_1000_pd[QC$mirror$band == "-7..-1"] /
  QC$mirror$rate_per_1000_pd[QC$mirror$band == "+1..+7"],
QC$mirror$rate_per_1000_pd[QC$mirror$band == "day 0"]))
```

::: {.callout-important}
## Limitations that must appear in any write-up

- KDIGO AKI requires a prior creatinine in the lookback window, so measured
  incidence is a joint function of injury and draw frequency. Initiators are
  drawn more often than non-initiators (see §1 and the `labdens` term in §4).
- **`Route` is not in this RPDR extract**, so inpatient prophylaxis cannot be
  fully separated from outpatient maintenance. Care setting is modelled in M12;
  route is not available without a new data request.
- No drug duration: exposure persists after a single order, and a substantial
  share of initiators have one record in follow-up (§12).
- Only `r QC$initiators$pct_followed_730`% of the cohort remains under
  creatinine surveillance at 730 days, so informative censoring is the dominant
  threat at this horizon — more so than at the 548 days used in the reference
  summaries.
- The parent cohort inherits `smoking` from the ICI-CKD master, which was
  derived with **no date restriction** and may include post-ICI diagnoses. It is
  adjusted for as `smoking_ever` and is not a baseline covariate.
:::

## Analyses deliberately not run

- **Excluding patients whose AKI preceded their PPI** — selection on a
  post-baseline outcome, which inflates the estimate.
- **Record count as dose-response** — AKI incidence rises with *future* record
  count in pre-exposure person-time, so it measures illness, not dose.
- **Censoring at `last_cre_days + 90`** — invents surveillance that did not occur.
- **Restricting the cohort to patients with a full 2 years of follow-up** — this
  would discard most of the cohort and select on survival, which is downstream
  of the exposure. Survival analysis does not require complete follow-up.

```{r}
#| label: session
#| echo: false
#| output: asis
cat("\n\n---\n\n*Built from `cache/qc.rds` created ",
    format(QC$built_at, "%Y-%m-%d %H:%M"),
    ". Aggregate results only; cells below ", CFG$SMALL_CELL, " suppressed.*\n", sep = "")
```
