---
title: "PPI → AKI after ICI — Analysis"
subtitle: "Time-varying Cox model, 730-day horizon"
author: "Tianqi Ouyang"
date: "`r Sys.Date()`"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
## About this page
Renders **aggregate results only** from `cache/*.rds` built by
[`data_management.qmd`](data_management.qmd). No patient-level row reaches the
output; cells below 11 are suppressed. Follow-up horizon is 730 days.
:::

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr); library(survival); library(broom); library(ggplot2); library(knitr)
})
source("R/ppi_functions.R")
have <- all(file.exists(file.path("cache", paste0(c("qc","pt","cp","cfg"), ".rds"))))
```

```{r}
#| label: load
#| include: false
if (!have) knitr::knit_exit()
QC  <- readRDS("cache/qc.rds")
pt  <- readRDS("cache/pt.rds")
cp  <- readRDS("cache/cp.rds")
CFG <- readRDS("cache/cfg.rds")
```

```{r}
#| label: header
#| echo: false
#| output: asis
c0 <- QC$cohort
cat(sprintf(paste0(
  "**%s** patients · **%s** start a PPI during follow-up · **%s** never do · ",
  "**%s** AKI events · %s person-years · median follow-up %s days.\n\n"),
  format(c0$n, big.mark=","), format(c0$ppi_initiators, big.mark=","),
  format(c0$never_ppi, big.mark=","), format(c0$aki_events, big.mark=","),
  format(c0$person_years, big.mark=","), c0$median_fu_days))
```

# 1 Cohort

```{r}
#| label: tbl-cohort
#| echo: false
kable(QC$ledger, caption = "How the 8,911 analytic cohort is reached.")
kable(QC$person_time,
      caption = "Person-time and crude AKI rate by time-varying PPI state.")
```

# 2 Time-varying Cox model

Every patient is unexposed at ICI start and becomes exposed only from the day
they fill a PPI, so their pre-PPI follow-up counts as unexposed person-time.
The model is `Surv(tstart, tstop, AKI) ~ PPI`.

```{r}
#| label: cox
m_crude <- coxph(Surv(tstart, tstop, event) ~ ppi, data = cp)

m_adj   <- coxph(Surv(tstart, tstop, event) ~ ppi +
                   age_ici + male + Race + dm + htn + eGFR_CRE_baseline, data = cp)

bind_rows(
  broom::tidy(m_crude, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "ppi") %>% mutate(model = "Crude"),
  broom::tidy(m_adj, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "ppi") %>% mutate(model = "Adjusted*")
) %>%
  transmute(Model = model,
            `HR for AKI (95% CI)` = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high),
            p = ifelse(p.value < .001, "<0.001", sprintf("%.3f", p.value))) %>%
  kable(caption = "Time-varying PPI and AKI. *Adjusted for age, sex, race, diabetes, hypertension, baseline eGFR.")
```

```{r}
#| label: ph-check
#| echo: false
z <- cox.zph(m_adj)
zt <- as.data.frame(z$table)["ppi", , drop = FALSE]
cat(sprintf("_Proportional-hazards check for the PPI term (cox.zph): p = %s._\n\n",
            ifelse(zt$p < .001, "<0.001", sprintf("%.3f", zt$p))))
```

# 3 Cumulative incidence of AKI

Cumulative incidence of AKI over two years for the whole cohort (death and end
of creatinine follow-up treated as censoring).

```{r}
#| label: fig-cuminc
#| echo: false
#| fig-height: 4.2
#| fig-cap: "Cumulative incidence of AKI within 730 days of ICI start (8,911 new-users)."
km <- survfit(Surv(fu_days, ev) ~ 1, data = pt)
data.frame(time = km$time, ci = 1 - km$surv) |>
  ggplot(aes(time, ci)) +
  geom_step(color = "#0B6E6D", linewidth = 0.9) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, NA)) +
  labs(x = "Days since ICI start", y = "Cumulative incidence of AKI") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: session
#| echo: false
#| output: asis
cat("\n---\n\n*Built from `cache/qc.rds` created ",
    format(QC$built_at, "%Y-%m-%d %H:%M"),
    ". Aggregate results only; cells below ", CFG$SMALL_CELL, " suppressed.*\n", sep = "")
```
