---
title: "PPI → AKI after ICI — Data Management"
subtitle: "New-user cohort, time-varying PPI exposure, 730-day horizon"
author: "Tianqi Ouyang"
date: "`r Sys.Date()`"
execute:
  eval: false
  warning: false
  message: false
---

::: {.callout-note}
## About this page

Executable specification of the pipeline. Chunks are not run at render time —
they read RPDR extracts that stay local. To run:

```r
knitr::purl("data_management.qmd", output = "cache/data_management.R", documentation = 0)
source("cache/data_management.R")
```

Helpers live in [`R/ppi_functions.R`](R/ppi_functions.R). Outputs land in
`cache/` (gitignored); the aggregate `cache/qc.rds` is what
[`analysis.qmd`](analysis.qmd) renders from.
:::

The analysis has four pieces:

1. **Cohort** — the 8,911 patients with no PPI at or before ICI start and at
   least one creatinine afterwards.
2. **Outcome** — AKI, computed from each patient's serial creatinine **after**
   the ICI start date.
3. **Exposure** — PPI, **time-varying**. Every patient is unexposed at ICI start
   (day 0); a patient who fills a PPI on day *k* becomes exposed from day *k*.
4. **Follow-up** — to the earliest of AKI, last creatinine, death, or 730 days.

---

# 1 Setup

```{r}
#| label: libraries
suppressPackageStartupMessages({
  library(readxl); library(dplyr); library(tidyr); library(data.table); library(survival)
})
source("R/ppi_functions.R")
```

```{r}
#| label: config
CFG <- list(
  master_xlsx = "/Users/to909/Desktop/Meg/Kavita ICI CKD/Final cohort/final_master_0114_raw.xlsx",
  rpdr_dir    = "/Users/to909/Partners HealthCare Dropbox/Tianqi Ouyang/CKD after ICI/RPDR",
  rpdr_stem   = "mes98_093025104659807332",
  HORIZON     = 730,          # days -- 2 years
  SMALL_CELL  = 11,
  cache       = "cache",
  REBUILD_RAW = FALSE
)
dir.create(CFG$cache, showWarnings = FALSE, recursive = TRUE)
rpdr  <- function(k) file.path(CFG$rpdr_dir, paste0(CFG$rpdr_stem, "_", k, ".txt"))
cpath <- function(x) file.path(CFG$cache, paste0(x, ".rds"))
cached <- function(name, expr) {
  p <- cpath(name)
  if (!CFG$REBUILD_RAW && file.exists(p)) return(readRDS(p))
  v <- force(expr); saveRDS(v, p); v
}
QC <- list()   # aggregate-only; rendered
```

---

# 2 Cohort (8,911)

Start from the ICI-CKD analytic master. Keep only new-users (no PPI on or before
ICI) with at least one post-ICI creatinine so AKI is ascertainable.

```{r}
#| label: load-master
master <- read_excel(CFG$master_xlsx) %>%
  standardize_colnames() %>%
  mutate(across(c(ICI_Dose_Date, Date_of_Birth, Date_Of_Death), to_date)) %>%
  distinct(EMPI, .keep_all = TRUE)

baseline <- master %>%
  transmute(
    EMPI, ICI_Dose_Date, Date_Of_Death,
    age_ici, male,
    Race = case_when(Race1 == "White" ~ "White", Race1 == "Black" ~ "Black",
                     Race1 == "Asian" ~ "Asian", TRUE ~ "Other/Unknown"),
    dm, htn, eGFR_CRE_baseline
  )
```

```{r}
#| label: ppi-exposure
## PPI dispensings. `<=` (baseline) and `>` (initiation) are complementary, so a
## new-user can never also be counted as an initiator.
med <- cached("med_ppi", {
  m <- data.table::fread(rpdr("Med"), sep = "|", quote = "", fill = TRUE,
                         showProgress = FALSE,
                         select = c("EMPI", "Medication_Date", "Medication"))
  m[, Medication_Date := to_date(Medication_Date, "%m/%d/%Y")]
  m[, EMPI := as.character(EMPI)]
  m[grepl(name_regex(PPI_NAMES), Medication, ignore.case = TRUE)][]
})
master_med <- med %>% inner_join(baseline %>% select(EMPI, ICI_Dose_Date), by = "EMPI")

ppi_pre  <- med_baseline(master_med, PPI_NAMES)                       # PPI on/before ICI
ppi_post <- med_first_after(master_med, PPI_NAMES) %>%               # first PPI after ICI
  rename(ppi_first_post_date = first_date)
```

```{r}
#| label: creatinine
## Serial creatinine (post-ICI creatinine feeds the AKI outcome; the last
## creatinine at any time defines end of surveillance).
cre <- cached("cre", {
  lab <- data.table::fread(rpdr("Lab"), sep = "|", quote = "", fill = TRUE,
                           showProgress = FALSE,
                           select = c("EMPI", "Seq_Date_Time", "Group_Id", "Result"))
  lab <- lab[Group_Id == "CRE"]
  lab[, `:=`(EMPI = as.character(EMPI),
             cre_date = to_date(Seq_Date_Time, "%m/%d/%Y"),
             cre = parse_lab_numeric(Result))]
  lab[!is.na(cre_date) & !is.na(cre), .(EMPI, cre_date, cre)][]
})

last_cre <- cre %>% group_by(EMPI) %>%
  summarise(last_cre_date = max(cre_date), .groups = "drop")
```

```{r}
#| label: cohort-ledger
ledger <- list(); step <- function(lbl, df) {
  ledger[[length(ledger) + 1]] <<- tibble(step = lbl, n = nrow(df)); df }

coh <- baseline %>% step("ICI-CKD analytic master (as saved)", .) %>%
  filter(is.na(Date_Of_Death) | Date_Of_Death >= ICI_Dose_Date) %>%
    step("- death recorded before ICI start", .) %>%
  anti_join(ppi_pre, by = "EMPI") %>%
    step("- PPI on or before ICI (keep new-users)", .) %>%
  left_join(last_cre, by = "EMPI") %>%
  mutate(last_cre_days = as.numeric(last_cre_date - ICI_Dose_Date)) %>%
  filter(!is.na(last_cre_days), last_cre_days > 0) %>%
    step("- no creatinine after ICI (keep ascertainable)", .)

QC$ledger <- bind_rows(ledger) %>% mutate(removed = lag(n) - n)
stopifnot(nrow(coh) == 8911)          # the analytic cohort
```

---

# 3 AKI outcome (post-ICI creatinine)

```{r}
#| label: aki
aki <- create_aki(cre, coh %>% select(EMPI, ICI_Dose_Date))

coh <- coh %>% left_join(aki %>% select(EMPI, aki, aki_date, aki_days), by = "EMPI")
```

---

# 4 Follow-up & time-varying exposure

```{r}
#| label: analysis-frame
pt <- coh %>%
  left_join(ppi_post, by = "EMPI") %>%
  mutate(
    ppi_day    = as.numeric(ppi_first_post_date - ICI_Dose_Date),
    death_days = as.numeric(Date_Of_Death - ICI_Dose_Date),
    ## Follow-up ends at the earliest of AKI, last creatinine, death, horizon.
    fu_days    = pmin(aki_days, last_cre_days, death_days, CFG$HORIZON, na.rm = TRUE),
    ## Event = AKI within follow-up.
    ev         = as.integer(!is.na(aki_days) & aki_days <= fu_days),
    ## Initiator = starts a PPI while still at risk.
    ppi_initiator = as.integer(!is.na(ppi_day) & ppi_day > 0 & ppi_day < fu_days)
  ) %>%
  filter(fu_days > 0)

## Impute the handful of missing baseline eGFR at the patient level (an
## adjusted model would otherwise silently drop those rows).
if (any(is.na(pt$eGFR_CRE_baseline)))
  pt$eGFR_CRE_baseline[is.na(pt$eGFR_CRE_baseline)] <-
    median(pt$eGFR_CRE_baseline, na.rm = TRUE)

QC$cohort <- tibble(
  n            = nrow(pt),
  ppi_initiators = sum(pt$ppi_initiator),
  never_ppi    = sum(pt$ppi_initiator == 0),
  aki_events   = sum(pt$ev),
  person_years = round(sum(pt$fu_days) / 365.25, 1),
  median_fu_days = median(pt$fu_days)
)
```

```{r}
#| label: counting-process
## One row per exposure interval: unexposed (0 -> ppi_day) then exposed
## (ppi_day -> end). Surv(tstart, tstop, event) ~ ppi.
cp <- build_cp_ppi(
  pt %>% transmute(id = EMPI, fu_days, ev, ppi_day,
                   age_ici, male, Race, dm, htn, eGFR_CRE_baseline)
)

## Person-time and crude rate by exposure state.
QC$person_time <- cp %>%
  mutate(days = tstop - tstart) %>%
  group_by(exposure = ifelse(ppi == 1, "On PPI (after initiation)", "Not on PPI")) %>%
  summarise(person_years = round(sum(days) / 365.25, 0),
            aki_events   = sum(event), .groups = "drop") %>%
  mutate(rate_per_100py = round(100 * aki_events / person_years, 1))

## Sanity: exactly the cohort's patients and events survive the split.
stopifnot(dplyr::n_distinct(cp$id) == nrow(pt), sum(cp$event) == sum(pt$ev))
```

---

# 5 Save

```{r}
#| label: save
saveRDS(pt,  cpath("pt"))    # patient-level (holds EMPI: local only)
saveRDS(cp,  cpath("cp"))    # counting-process frame for the Cox model
saveRDS(CFG, cpath("cfg"))

walk_ok <- vapply(QC, function(x) if (is.data.frame(x)) { assert_no_phi(x); TRUE } else TRUE,
                  logical(1))
QC$built_at <- Sys.time()
saveRDS(QC, cpath("qc"))
```

---

# 6 Rendered summary

```{r}
#| label: qc-render
#| eval: true
#| echo: false
qc_path <- file.path("cache", "qc.rds")
if (!file.exists(qc_path)) {
  cat("_Pipeline has not been run yet — see the callout at the top._")
  knitr::knit_exit()
}
QC <- readRDS(qc_path)
knitr::kable(QC$ledger,       caption = "Cohort ledger (each step asserted in code)")
knitr::kable(t(QC$cohort),    caption = "Analytic cohort")
knitr::kable(QC$person_time,  caption = "Person-time and crude AKI rate by time-varying PPI state")
```
