---
title: "PPI → AKI after ICI — Co-timing Diagnostics"
subtitle: "Is the association a drug effect, or the encounter that produced the prescription?"
author: "Tianqi Ouyang"
date: "`r Sys.Date()`"
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
## About this page
Renders **aggregate results only** from `cache/*.rds` built by
[`data_management.qmd`](data_management.qmd). No patient-level row reaches the
output; cells below 11 are suppressed. Follow-up horizon is 730 days.
:::

The time-varying Cox model in [`analysis.qmd`](analysis.qmd) answers one
question: *is the AKI hazard higher while a patient is on a PPI?* It cannot
answer a second one: *is the PPI doing it?* A drug ordered at the same clinical
moment as the outcome will look causal to any model that only knows exposure
status at each instant.

Three diagnostics separate the two. All of them re-zero the clock on each
patient's **own initiation date** rather than on the ICI infusion.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr); library(survival); library(broom)
  library(ggplot2); library(knitr)
})
source("R/ppi_functions.R")
have <- all(file.exists(file.path("cache", paste0(c("pt", "cfg"), ".rds"))))
```

```{r}
#| label: load
#| include: false
if (!have) knitr::knit_exit()
pt  <- readRDS("cache/pt.rds")
CFG <- readRDS("cache/cfg.rds")

covars <- c("age_ici", "male", "dm", "htn", "eGFR_CRE_baseline")
d <- pt %>% transmute(id = EMPI, fu_days, ev, ppi_day, ster_day,
                      age_ici, male, dm, htn, eGFR_CRE_baseline)
```

# 1 The mirror window

If a PPI causes AKI, the rate should be flat *before* the prescription and
raised *after* it. This table places windows on both sides of each initiator's
own PPI date. Person-time is counted over the observation window (last
creatinine, death, or horizon) — **not** the at-risk window, which stops at the
AKI and would delete every patient whose AKI came first.

```{r}
#| label: tbl-mirror
#| echo: false
mw <- mirror_window(pt, pt$obs_days)
mw %>%
  mutate(side = case_when(as.numeric(sub("\\.\\..*", "", window)) < 0 ~ "before",
                          window == "+0..+0" ~ "day of prescription",
                          TRUE ~ "after")) %>%
  transmute(Window = window, Side = side,
            `Person-days` = format(person_days, big.mark = ","),
            AKI = aki, `Rate / 1,000 person-days` = rate_per_1000_pd) %>%
  kable(caption = "AKI incidence in windows placed relative to each initiator's own first PPI.")
```

```{r}
#| label: fig-mirror
#| echo: false
#| fig-height: 4.2
#| fig-cap: "AKI rate per 1,000 person-days by window relative to the first PPI. A drug effect raises only the right-hand side; a shared cause is symmetric."
mw %>%
  mutate(window = factor(window, levels = window),
         side = ifelse(as.numeric(sub("\\.\\..*", "", window)) < 0, "Before PPI",
                ifelse(window == "+0..+0", "Day of PPI", "After PPI"))) %>%
  ggplot(aes(window, rate_per_1000_pd, fill = side)) +
  geom_col() +
  scale_fill_manual(values = c("Before PPI" = "#0B6E6D", "Day of PPI" = "#98301F",
                               "After PPI" = "#C8862A"), name = NULL) +
  labs(x = "Days relative to first PPI", y = "AKI per 1,000 person-days") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top")
```

::: {.callout-important}
## How to read this
The rate in the week *before* the prescription is close to the rate in the week
*after* it, and the single highest-rate window is the prescription day itself.
A drug cannot raise risk in the week before it is taken, so a symmetric peak
centred on day 0 points to a shared cause — the acute admission that produces
both the stress-ulcer prophylaxis order and the daily creatinine draws.
:::

# 2 Hazard by time since initiation

Splitting the exposed person-time by how long the patient has actually been on
the drug. A pharmacologic effect should persist with continued exposure; an
artifact of the ordering encounter should decay.

```{r}
#| label: tbl-bands
#| echo: false
cp_b <- build_cp_bands(d)

## Structural checks — the split must preserve patients, events and person-time.
stopifnot(dplyr::n_distinct(cp_b$id) == nrow(d),
          sum(cp_b$event) == sum(d$ev),
          abs(sum(cp_b$tstop - cp_b$tstart) - sum(d$fu_days)) < 1,
          !any(is.na(cp_b$band)))

m_band <- coxph(as.formula(paste("Surv(tstart, tstop, event) ~ band + ster +",
                                 paste(covars, collapse = " + "))), data = cp_b)

rate <- cp_b %>% group_by(band) %>%
  summarise(person_years = round(sum(tstop - tstart) / 365.25, 1),
            aki = sum(event), .groups = "drop") %>%
  mutate(rate_per_100py = round(100 * aki / person_years, 1))

hr <- tidy(m_band, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(grepl("^band", term)) %>%
  transmute(band = sub("^band", "", term),
            HR = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high))

rate %>% left_join(hr, by = "band") %>%
  mutate(band = as.character(band),
         HR = ifelse(is.na(HR), "1.00 (reference)", HR)) %>%
  rename(`Time since first PPI` = band, `Person-years` = person_years,
         AKI = aki, `Rate / 100 py` = rate_per_100py, `HR (95% CI)` = HR) %>%
  kable(caption = "Adjusted for age, sex, diabetes, hypertension, baseline eGFR, and time-varying systemic steroids.")
```

# 3 Induction-period ladder

Credit exposure only from *lag* days after the prescription, and drop the
induction window from the risk set entirely rather than handing it back to the
comparator. A real effect plateaus; a co-timing artifact decays monotonically.

```{r}
#| label: tbl-ladder
#| echo: false
induction_ladder(d, lags = c(0, 30, 90, 180), rhs = c("ster", covars)) %>%
  transmute(`Induction period (days)` = induction_days,
            `AKI events` = events,
            `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, lo, hi)) %>%
  kable(caption = "PPI effect on AKI as the prescription is separated from the risk window.")
```

# 4 Death as a negative-control outcome

The same exposure structure, with **death** as the outcome. Creatinine sampling
can manufacture an AKI; nothing manufactures a death. Follow-up here runs to
death or the horizon and is *not* censored at the last creatinine, because death
is ascertained from the registry rather than from labs.

```{r}
#| label: tbl-death
#| echo: false
d_death <- pt %>%
  transmute(id = EMPI, ppi_day, ster_day, age_ici, male, dm, htn, eGFR_CRE_baseline,
            fu_days = pmin(death_days, CFG$HORIZON, na.rm = TRUE),
            ev      = as.integer(!is.na(death_days) & death_days <= CFG$HORIZON)) %>%
  filter(fu_days > 0)

cp_d <- build_cp_bands(d_death)
m_d  <- coxph(as.formula(paste("Surv(tstart, tstop, event) ~ band + ster +",
                               paste(covars, collapse = " + "))), data = cp_d)

bind_rows(
  tidy(m_band, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("^band", term)) %>% mutate(outcome = "AKI"),
  tidy(m_d, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(grepl("^band", term)) %>% mutate(outcome = "Death")
) %>%
  transmute(band = sub("^band", "", term), outcome,
            hr = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high)) %>%
  tidyr::pivot_wider(names_from = outcome, values_from = hr) %>%
  rename(`Time since first PPI` = band, `AKI HR (95% CI)` = AKI,
         `Death HR (95% CI)` = Death) %>%
  kable(caption = "Same cohort, same time-varying exposure, two outcomes.")
```

::: {.callout-important}
## How to read this
If starting a PPI predicts **death** at least as strongly as it predicts kidney
injury, then whatever the exposure is marking is general illness severity, not
nephrotoxicity.
:::

# 5 Steroid co-initiation

The imbalance that the time-varying steroid term exists to absorb.

```{r}
#| label: tbl-steroid
#| echo: false
QC <- readRDS("cache/qc.rds")
QC$steroid %>%
  rename(Group = group, N = n, `Started a systemic steroid` = started_steroid, `%` = pct) %>%
  kable(caption = "Systemic steroid initiation after ICI start, by whether the patient ever starts a PPI.")
```

```{r}
#| label: session
#| echo: false
#| output: asis
cat("\n---\n\n*Aggregate results only; cells below ", CFG$SMALL_CELL,
    " suppressed. Horizon ", CFG$HORIZON, " days.*\n", sep = "")
```
